---
title: "Getting started with odelia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odelia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

This vignette walks through the minimal setup for solving an ODE system with
`odelia`, using the classic
[Lorenz system](https://en.wikipedia.org/wiki/Lorenz_system) as the worked
example. The Lorenz system is built into the package, so everything below runs
against `library(odelia)` directly — no C++ compilation needed.

```{r setup}
library(odelia)
```

## The core abstractions

A simulation in `odelia` is assembled from a few pieces:

- a **system** — the ODE model itself (parameters, state, and a rule for the
  rates `dy/dt`);
- a **control** object (`OdeControl`) — the solver's tolerances and step sizes;
- a **solver** (or *runner*) — which drives the adaptive RK4-5 stepper forward
  over the times you request and collects the solution history.

## Define the system

The Lorenz system has three state variables and three parameters
(`sigma`, `R`, `b`). We create it and set an initial state. The Lorenz system is
autonomous (its rates do not depend on time directly), so no time is required.

```{r system}
lz <- LorenzSystem$new(sigma = 10, R = 28, b = 8 / 3)

# Inspect stored parameters
lz$pars()

# Set the initial state (x, y, z)
lz$set_state(c(1, 1, 1))

# We can query the current state and rates directly
lz$state()
lz$rates()
```

## Solve the system

Create a control object (the defaults are sensible) and a solver, then advance
with adaptive time stepping over the times we want output at.

```{r solve}
ctrl <- OdeControl$new()

runner <- Lorenz_Solver$new(lz$ptr, ctrl$ptr)
runner$advance_adaptive(seq(0, 100, by = 0.01))

out <- runner$history()
head(out)
```

`history()` returns a tibble with one row per requested output time. Because the
stepper is adaptive, the solver may take more internal steps than there are
output rows — the internal evaluation times are available via `runner$times()`.

## Plot the attractor

```{r plot}
plot(out$x, out$z,
  type = "l",
  xlab = "x", ylab = "z",
  main = "Lorenz attractor (via odelia)"
)
```

## Controlling the solver

`OdeControl` exposes the solver's tuning knobs — absolute/relative tolerances,
state and derivative scaling, and minimum/maximum/initial step sizes. You can
inspect the current settings or tighten them:

```{r control}
ctrl$get_controls()

# Tighten tolerances for a more accurate (but slower) solve
ctrl$set_tol_abs(1e-8)
ctrl$set_tol_rel(1e-8)
```

## Where next

- For models with **external time-varying forcing**, and how to define your own
  ODE system in C++, see
  [Building your own model with external drivers](https://traitecoevo.github.io/odelia/articles/leaf-thermal.html).
- To recover parameters from data using exact gradients, see
  `vignette("parameter-fitting", package = "odelia")`.
