---
title: "Getting started with odelia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odelia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

This vignette walks through the minimal setup for solving an ODE system with
`odelia`, using the classic
[Lorenz system](https://en.wikipedia.org/wiki/Lorenz_system) as the worked
example. The Lorenz system is built into the package, so everything below runs
against `library(odelia)` directly — no C++ compilation needed.

```{r setup}
library(odelia)
```

## The core abstractions

A simulation in `odelia` is assembled from a few pieces:

- a **system** — the ODE model itself (parameters, state, and a rule for the
  rates `dy/dt`);
- a **control** object (`OdeControl`) — the solver's tolerances and step sizes;
- a **solver** (or *runner*) — which drives the adaptive RK4-5 stepper forward
  over the times you request and collects the solution history.

## Define the system

The Lorenz system has three state variables and three parameters
(`sigma`, `R`, `b`). We create it and set an initial state. The Lorenz system is
autonomous (its rates do not depend on time directly), so no time is required.

```{r system}
lz <- LorenzSystem$new(sigma = 10, R = 28, b = 8 / 3)

# Inspect stored parameters
lz$pars()

# Set the initial state (x, y, z)
lz$set_state(c(1, 1, 1))

# We can query the current state and rates directly
lz$state()
lz$rates()
```

## Solve the system

Create a control object (the defaults are sensible) and a solver, then advance
with adaptive time stepping over the times we want output at.

```{r solve}
ctrl <- OdeControl$new()

runner <- Lorenz_Solver$new(lz$ptr, ctrl$ptr)
runner$advance_adaptive(seq(0, 100, by = 0.01))

out <- runner$history()
head(out)
```

`history()` returns a tibble with one row per requested output time. Because the
stepper is adaptive, the solver may take more internal steps than there are
output rows — the internal evaluation times are available via `runner$times()`.

## Plot the attractor

```{r plot}
plot(out$x, out$z,
  type = "l",
  xlab = "x", ylab = "z",
  main = "Lorenz attractor (via odelia)"
)
```

## Controlling the solver

`OdeControl` exposes the solver's tuning knobs — absolute/relative tolerances,
state and derivative scaling, and minimum/maximum/initial step sizes. You can
inspect the current settings or tighten them:

```{r control}
ctrl$get_controls()

# Tighten tolerances for a more accurate (but slower) solve
ctrl$set_tol_abs(1e-8)
ctrl$set_tol_rel(1e-8)
```

## Choosing an integration method

`odelia` ships three steppers, chosen with the `method` argument when you create
a solver:

- `"rkck"` (the default for a compiled system) — an explicit adaptive Cash-Karp
  Runge-Kutta 4(5) method. Fast and accurate for **non-stiff** problems like
  Lorenz.
- `"dopri"` — the explicit Dormand-Prince 5(4) pair, the method behind
  `deSolve`'s `ode45`, with a **dense output** of its own order: the state at
  any time inside a step, from the step's own stages, at no extra evaluation.
  The default for a right-hand side written in R (below).
- `"rodas"` — an implicit, adaptive RODAS4(3) Rosenbrock method for **stiff**
  problems. Each step forms the Jacobian once (by automatic differentiation for
  a compiled system, or from the system's own Jacobian hook) plus a handful of
  linear solves, so the step size is limited by accuracy rather than stability. On a stiff problem the explicit method is
  forced into tiny steps (or fails), while RODAS keeps large steps.

```{r method}
runner_stiff <- Lorenz_Solver$new(lz$ptr, ctrl$ptr, method = "rodas")
```

All three share the same adaptive step-size controller and `OdeControl`
tolerances, so you can switch between them without changing anything else.

To use `method = "rodas"` with your own C++ system, give it either a
`template<class U> System<U> rebind() const` method (a one-liner that copies the
system with its scalar type swapped) so its right-hand side can be
differentiated for the Jacobian, as the Lorenz example does, or an
`ode_jacobian(y, t, dydt, J)` method that fills the Jacobian itself -- the
header `ode_jacobian.hpp` has `fd_jacobian()` for a one-line finite-difference
version. The implicit stepper currently runs on the passive (non-AD) solver;
differentiating a fit *through* RODAS is planned (odelia issue #36).

## Solving a system written in R

Everything above drives a system compiled in C++. From 0.6.0 the same solver,
with the same controller and both steppers, also takes a right-hand side
written in R. There is nothing to compile: `ode_solve()` is shaped like
`deSolve::ode()`, so a function written for deSolve runs unchanged, and returns
a matrix with a `time` column and one column per state variable.

```{r r-rhs}
lorenz_r <- function(t, y, p) {
  list(c(p[["sigma"]] * (y[2] - y[1]),
         y[1] * (p[["R"]] - y[3]) - y[2],
         y[1] * y[2] - p[["b"]] * y[3]))
}
pars <- c(sigma = 10, R = 28, b = 8 / 3)

out_r <- ode_solve(lorenz_r, y0 = c(x = 1, y = 1, z = 1),
                   times = seq(0, 2, by = 0.1), parms = pars,
                   autonomous = TRUE)
head(out_r)
```

The steps are the controller's own, and the rows asked for are read off each
step's dense output (`dense = TRUE`, the default), so asking for ten thousand
rows costs the same integration as asking for ten. With `dense = FALSE` a step
is made to end on every requested time instead, which is what `deSolve`'s
`ode45` does and what makes a fine output grid expensive there.

The implicit stepper needs the Jacobian of the right-hand side. Supply it as
`jacfunc(t, y, parms)`, an n by n matrix whose column j is the derivative with
respect to `y[j]`, or leave it out and it is formed by forward differences at
`n` extra evaluations per step. `counts()` says what the solve cost, which for a
right-hand side that is itself expensive is the only number that matters:

```{r r-rhs-rodas}
lorenz_jac <- function(t, y, p) {
  matrix(c(-p[["sigma"]], p[["R"]] - y[3], y[2],
           p[["sigma"]], -1, y[1],
           0, -y[1], -p[["b"]]), 3, 3)
}
out_rodas <- ode_solve(lorenz_r, y0 = c(x = 1, y = 1, z = 1),
                       times = seq(0, 2, by = 0.1), parms = pars,
                       jacfunc = lorenz_jac, method = "rodas", autonomous = TRUE)
counts(out_rodas)
```

A stiff problem is where the implicit stepper earns its keep. Van der Pol with
a small `eps` forces the explicit stepper into thousands of tiny steps for
stability while RODAS stays accuracy-limited:

```{r r-rhs-stiff}
vdp <- function(t, y, eps) {
  list(c(y[2], ((1 - y[1]^2) * y[2] - y[1]) / eps))
}
times <- seq(0, 2, by = 0.2)
stiff_rodas <- ode_solve(vdp, c(2, 0), times, parms = 1e-4, method = "rodas", autonomous = TRUE)
stiff_dopri <- ode_solve(vdp, c(2, 0), times, parms = 1e-4, method = "dopri", autonomous = TRUE)
rbind(rodas = counts(stiff_rodas), dopri = counts(stiff_dopri))
```

Underneath `ode_solve()` is the R6 class `OdeSolver`, which is driven a step
at a time. That is what a caller with *events* needs: after each accepted step
it can look at the state, change it, and even change its length, before the
next step. The right-hand side here is `rhs(t, y)` returning the rates
directly, and a callback that finds its state impossible calls
`domain_error()` to have the step rejected and retried smaller rather than
committed:

```{r r-rhs-stepwise}
logistic <- function(t, y) {
  if (y[1] < 0 || y[1] > 1) domain_error("y has left [0, 1]")
  50 * y[1] * (1 - y[1])
}
s <- OdeSolver$new(logistic, y0 = 0.5, autonomous = TRUE)
while (s$time() < 1) {
  s$step(time_max = 1)
}
c(time = s$time(), y = s$state())
s$counts()
```

Between steps `s$set_state(y, time)` re-seeds the state at any length and
`s$set_step_size(h)` carries a step size across it; `s$advance_collect(times)`
is the output loop of `ode_solve()`, a matrix of states at the times asked for.

Each stage of a step is one call into R, so for a cheap right-hand side this
runs at the speed of the R function and the call into it, which is the speed
`deSolve` runs an R function at too (on Lorenz, 0.95 µs a call here against
1.03 µs there, with the R function itself 0.6 µs of each). What makes odelia
faster on a fine output grid is the dense output: `deSolve::ode45` lands a step
on every requested time, odelia does not have to. Lorenz to t = 100 at 1e-6
with 10001 rows takes 27 ms here against 62 ms for `ode45` and 33 ms for
`lsoda`, with the same R function and the same accuracy against a tight
reference. The compiled route above is a hundred times faster still, because
nothing crosses into R per stage. For a right-hand side that is itself a large
computation the solver's own cost is nothing, and what `counts()` reports is
the whole cost of the integration.

## Where next

- For models with **external time-varying forcing**, and how to define your own
  ODE system in C++, see
  [Building your own model with external drivers](https://traitecoevo.github.io/odelia/articles/leaf-thermal.html).
- To recover parameters from data using exact gradients, see
  `vignette("parameter-fitting", package = "odelia")`.
