---
title: "Building your own model with external drivers"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

The [getting-started vignette](../articles/odelia.html) uses the Lorenz system,
which ships compiled into the package. This article shows the other half of the
story: **how to define your own ODE system in C++ and drive it with external,
time-varying forcing.** It is the workflow you would follow to use `odelia` in
your own project.

We use a simple *leaf thermal model* — a single ODE for leaf temperature that is
forced by air temperature (an external driver) and cooled by transpiration. The
complete, copy-able sources live in the installed package under
`system.file("examples/leaf_thermal", package = "odelia")`.

> This is a website-only article rather than a packaged vignette, because it
> compiles model-specific C++ at build time. That keeps the package's
> cross-platform `R CMD check` fast and toolchain-light.

## Anatomy of an `odelia` model

A model that uses `odelia` is assembled from four files:

| File | Role |
|------|------|
| `src/leaf_thermal_system.hpp`     | The **system**: a C++ class defining the ODE. |
| `src/leaf_thermal_interface.cpp`  | An **Rcpp interface** exposing the system to R. |
| `R/leaf_thermal_interface.R`      | **R6 wrappers** giving a friendly R API. |
| this article                      | A runnable demonstration. |

### The system contract

A system is a class templated on its scalar type `T` (so the same code works
with `double` and with the AD number type). It must implement a small contract:

```cpp
template <typename T = double>
class LeafThermalSystem {
public:
  // number of state variables
  size_t ode_size() const { return ode_dimension; }

  // read state in from an iterator (and refresh drivers/rates)
  template <typename Iterator>
  Iterator set_ode_state(Iterator it, double time_);

  // write the current state out to an iterator
  template <typename Iterator>
  Iterator ode_state(Iterator it) const;

  // write the current rates (dy/dt) out to an iterator
  template <typename Iterator>
  Iterator ode_rates(Iterator it) const;
};
```

Templating on `T` is what makes the system **AD-ready** — see the
[parameter-fitting vignette](../articles/parameter-fitting.html).

### Drivers

External forcing enters through a `drivers::Drivers` object. The system holds a
reference to it and queries the current value on each step:

```cpp
void initialize_drivers(const drivers::Drivers &drv);   // attach drivers
void set_drivers();                                     // query at current time
```

On the R side, `Drivers` fits a smooth cubic-spline interpolation through a time
series, so the system can be evaluated at any time the adaptive stepper lands on.
A `Drivers` object can hold several variables and accepts any time grid that
covers the simulation.

## Compile the model

Compiling a model interface requires the `odelia` headers on the include path.
Loading `odelia` exposes its compiled symbols (including the XAD runtime)
globally via the package `.onLoad`, so the temporary shared object that
`Rcpp::sourceCpp()` builds can resolve them — no manual `odelia_load_dll()`
call is needed.

```{r compile}
library(odelia)

ex_dir <- system.file("examples/leaf_thermal", package = "odelia")

# Put the odelia headers on the include path for sourceCpp
Sys.setenv(PKG_CPPFLAGS = paste0("-I", system.file("include", package = "odelia")))

# Compile the model-specific interface, then load the R6 wrappers
Rcpp::sourceCpp(file.path(ex_dir, "src", "leaf_thermal_interface.cpp"), verbose = FALSE)
source(file.path(ex_dir, "R", "leaf_thermal_interface.R"))
```

## Define drivers and run

Here the air-temperature driver is a sinusoidal daily cycle over two days.

```{r run}
# A daily temperature cycle
p <- list(Tmean = 32, Tamp = 6, tpeak = 15)
time_driver <- seq(0, 48, by = 0.25)
t_air <- p$Tmean + p$Tamp * sin(2 * pi * (time_driver - p$tpeak) / 24)

drivers <- Drivers$new()
drivers$set_variable("temperature", time_driver, t_air)

# Build the system with its parameters and drivers
pars <- LeafThermalSystemPars()
lz <- LeafThermalSystem$new(pars, drivers)
lz$set_state(c(25), 0) # initial leaf temperature, initial time

# Solve
ctrl <- OdeControl$new()
runner <- LeafThermalSolver$new(lz$ptr, ctrl$ptr, drivers$ptr)
times <- seq(0, 48, by = 0.5)
runner$advance_adaptive(times)

out <- runner$history()
out$time <- times
head(out)
```

```{r plot}
plot(out$time, out$T_air,
  type = "l", col = "grey50", lwd = 2,
  xlab = "time (h)", ylab = "temperature (°C)",
  ylim = range(out$T_air, out$T_LC)
)
lines(out$time, out$T_LC, col = "firebrick", lwd = 2)
legend("topright",
  legend = c("air", "leaf"),
  col = c("grey50", "firebrick"), lwd = 2, bty = "n"
)
```

The leaf temperature tracks the air temperature but is buffered by
transpirational cooling — exactly the behaviour we expect.

## Comparing strategies

Because building and running a system is cheap, comparing parameterisations is
just a loop. Here we double the maximum transpiration conductance and re-run:

```{r strategies}
run_strategy <- function(g_tr_max) {
  pars <- LeafThermalSystemPars()
  pars$g_tr_max <- g_tr_max
  lz <- LeafThermalSystem$new(pars, drivers)
  lz$set_state(c(25), 0)
  runner <- LeafThermalSolver$new(lz$ptr, ctrl$ptr, drivers$ptr)
  runner$advance_adaptive(times)
  runner$history()$T_LC
}

base_g <- LeafThermalSystemPars()$g_tr_max
T_baseline <- run_strategy(base_g)
T_high_cond <- run_strategy(2 * base_g)

plot(times, out$T_air,
  type = "l", col = "grey70", lwd = 2,
  xlab = "time (h)", ylab = "leaf temperature (°C)",
  ylim = range(out$T_air, T_baseline, T_high_cond)
)
lines(times, T_baseline, col = "firebrick", lwd = 2)
lines(times, T_high_cond, col = "steelblue", lwd = 2)
legend("topright",
  legend = c("air", "baseline", "2x conductance"),
  col = c("grey70", "firebrick", "steelblue"), lwd = 2, bty = "n"
)
```

Higher conductance means stronger transpirational cooling, so the leaf stays
closer to — or below — air temperature at the hottest part of the day.

## Where next

- The full, copy-able sources are in
  `system.file("examples/leaf_thermal", package = "odelia")` — start there when
  building your own model.
- To calibrate a model's parameters against data using exact gradients, see the
  [parameter-fitting vignette](../articles/parameter-fitting.html).
