---
title: "Reverse mode: one solve, every parameter"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Calibrating a model against data needs the derivative of a solved trajectory with respect to every parameter. Forward mode costs one solve per parameter. Reverse mode costs one solve for all of them, but it must first record every operation the solve performed, and a run of several thousand adaptive steps performs far more of them than memory holds. This article is how `odelia` does it, from C++, and what a System has to provide to take part. The R surface, `$set_target()` and `$fit()`, is the last section.

> This is a website-only article rather than a packaged vignette, because it compiles C++ at build time.

## The whole of it in eleven lines

```{r setup}
library(odelia)
# The same flags the package is built with. The second pair matters: XAD reaches
# its tape through a variable whose storage class these set, so a translation
# unit built without them reaches a different tape.
Sys.setenv(PKG_CPPFLAGS = paste0(
  "-I", shQuote(system.file("include", package = "odelia")),
  " -DXAD_NO_THREADLOCAL -DXAD_USE_STRONG_INLINE"))
```

```{Rcpp lorenz-adjoint}
// [[Rcpp::plugins(cpp20)]]
#include <Rcpp.h>
#include <odelia/ode_solver.hpp>
#include <odelia/adjoint.hpp>
#include <examples/lorenz_system.hpp>

using namespace odelia::ode;

// d x(t_end) / d (sigma, R, b) and d x(t_end) / d (x, y, z)(0), from one solve.
// [[Rcpp::export]]
Rcpp::List lorenz_adjoint(double sigma, double R, double b, double t_end) {
  LorenzSystem<double> system(sigma, R, b);
  Solver<LorenzSystem<double>> s(system, OdeControl());
  s.set_keep_states(true);                        // keep one step_record per step
  s.advance_adaptive(std::vector<double>{0.0, t_end});   // the ordinary solve

  adjoint_rows lambda = adjoint_rows::one_row({1.0, 0.0, 0.0});  // seed: d x(t_end)
  adjoint_rows dp(1, 3);                          // one row per seed, zeroed
  s.solve_adjoint(lambda, dp);                    // one pass, every parameter

  return Rcpp::List::create(
    Rcpp::Named("x")         = s.state()[0],
    Rcpp::Named("steps")     = static_cast<int>(s.recording().size()) - 1,
    Rcpp::Named("d_params")  = std::vector<double>(dp[0].begin(), dp[0].end()),
    Rcpp::Named("d_initial") = std::vector<double>(lambda[0].begin(), lambda[0].end()));
}
```

Three lines are new against an ordinary solve: `set_keep_states`, the seed, and `solve_adjoint`. The seed says which output is wanted, as a vector over the final state; here the first component. `dp` comes back holding that output's derivative with respect to each parameter in the order the System's `ad_parameters()` lists them, and `lambda` is replaced by the derivative with respect to the initial state.

```{r run}
r <- lorenz_adjoint(10, 28, 8 / 3, 2)
r
```

The answer is refereed here against a central difference of the same solve. The two agree to the integrator's own tolerance, not to rounding: perturbing a parameter also moves the adaptive step sequence, so the difference is of a slightly different discretisation. The sweep differentiates the run that was taken, with its steps held fixed.

```{r fd}
fd <- sapply(1:3, function(j) {
  h <- 1e-5
  p <- c(10, 28, 8 / 3)
  up <- p; up[j] <- up[j] + h
  dn <- p; dn[j] <- dn[j] - h
  (lorenz_adjoint(up[1], up[2], up[3], 2)$x - lorenz_adjoint(dn[1], dn[2], dn[3], 2)$x) / (2 * h)
})
rbind(sweep = r$d_params, difference = fd, relative = abs(r$d_params - fd) / abs(fd))
```

The cost of `solve_adjoint` is one more pass over the trajectory, which re-runs the model's rate evaluations as it goes. It is not one pass per parameter, which is the point, and it is not free.

## Forward, then reverse

```
  FORWARD                              REVERSE
  advance_adaptive(times)              solve_adjoint(lambda, dp)
    |                                    |
    +-> step()                           +-> range N ... 1          widest first
    |     6 x ode_rates()                |     |
    |     error, accept or shrink        |     +-> step k ... 1      last to first
    |                                    |     |     +- load rec[k-1].state
    +-> push_step() --> rec[k] ----------+     |     +- re-run the 6 stages ----+
                                               |     +- sweep, once per seed    |
                                               |     +- clear the tape          |
                                               |                                |
                                               +-> at an insertion:             |
                                                     sweep apply_insertion      |
                                                                                |
        System::ode_rates <-----------------------------------------------------+
        runs twice per step: forward at double, again here at the adjoint scalar
```

The forward pass stores one `step_record` per accepted step and nothing else. The reverse pass takes the rows last to first; for each it loads the state the step started from, re-runs the step's six stages on the adjoint scalar so they are recorded on the tape, sweeps that recording once per seed, and clears the tape. Peak tape is therefore one step's arithmetic at any run length. What accumulates across a run is the rows, which are cheap.

Because the reverse pass re-runs the model, anything the model caches between calls, or that depends on the order rates are computed in, differs between the two passes unless the System makes it agree. The `solved` channel below is how a System makes it agree for the one class of value that cannot be recomputed.

## What a recording holds

```
    step_record<System> : instruction
    |-- time        double         the time this step reached
    |-- step_size   double         h, as taken; NaN on the row the run started from
    |-- insertion   bool           did the state vector grow here
    |-- subdivided  bool           a pinned step crossed in several sub-steps
    |-- state[]     state_type     the state at `time`
    +-- solved[6]   solved_values  what its six rate evaluations solved for
```

Memory for a run of `n` steps on a state of width `w` is `n * (w + 6 * sizeof(solved_values))`. Stage rates and intermediate stage states are absent by choice: recomputing them costs six rate evaluations per step, holding them costs the whole trajectory, and for a right-hand side this cheap relative to its own length the recomputation wins.

`solved` exists for the one class of value that is neither cheap to recompute nor derivable from the state: anything a root-find produced inside a stage. For a plant model that is a leaf's operating point, found by iteration, whose value depends on where the iteration started. A later pass cannot re-derive it; it has to be told. A System that solves for nothing declares nothing and the slot is empty.

The row is six long, not five. Five entries are a step's stages; the sixth is the evaluation at the state the step ends at, which first-same-as-last hands the next step as its own first stage. A sweep reads the first five and re-derives the first stage at the state it was handed. A forward replay reads all six, because re-deriving is what a replay exists to avoid, and a step whose first stage was re-derived is wrong at first order in `h`.

A pinned step (`advance_fixed`) that the System's domain refused is crossed in several sub-steps and recorded as one row of the interval, holding only the last sub-step's solved values. The row is marked `subdivided`, and a sweep or a replay refuses it by name rather than treating the interval as one step. Record a run adaptively, or pin it finely enough that nothing is refused.

Only `method = "rkck"` records a run. `"dopri"` and `"rodas"` keep no per-stage row and refuse a sweep or a replay.

## Where the state grows

A System may introduce new state entries on a schedule fixed before the run. The schedule is an input to the solve, so the time at which an insertion happens carries no derivative with respect to anything the solve computes, which is what makes the widening a linear map that can be transposed instead of a discontinuity that cannot. An insertion whose time depends on the parameters is a different map, and nothing here computes its adjoint.

Going forward the vector only grows, so going backward the sweep only narrows. The narrowing is not done by dropping entries. The System's own widening map, `apply_insertion`, is recorded and swept like any other function:

```cpp
  // forward:  y_wide = apply_insertion(time, y_narrow)
  state_and_parameter_adjoints(widened, rec[at - 1].state,
                               lambda,             // in:  adjoint at wide width
                               insert,             //      the map, as a lambda
                               narrowed,           // out: adjoint at narrow width
                               parameter_adjoint); //      accumulated here too
  lambda = std::move(narrowed);
```

Two things follow. The map runs on the adjoint scalar with the parameters active, so a newborn's initial conditions contribute parameter derivatives: the steps are not the only source of them. And the recording of the map is taken on a System at the wide width and cannot be swept at the narrow one, so the sweep rebinds the System once per range. A range is a run of rows at one width; a System of fixed width has one, and a System that widens opens a new one at each insertion. `solve_adjoint` returns how many it swept.

A caller can also ask for cuts at rows of its own choosing, with `extra_stops`, for a partial sweep. A cut is a row the sweep resumes at; an insertion is a row it carries the adjoint across, so it resumes one row below, on the state the map ran on. A row that is both is treated as the insertion.

The seed batch, `lambda`, is replaced as the sweep narrows. The parameter batch is accumulated with `+=` across ranges, so the caller zeroes it, and the two must be different objects; the same object for both is refused.

## A derivative obtained somewhere else

Inside a stage, `ode_rates` may reach a quantity a solve found by iteration. Recording the iterations would differentiate the solver: the answer becomes the sensitivity of wherever that particular sequence of iterations stopped, which depends on the starting guess and the tolerance. `record_with_derivatives(value, rows, into)` in `implicit_node.hpp` puts the number on the tape carrying rows it was handed instead:

```
    out = value + sum_i d_i * (x_i - to_passive(x_i))
                        \_______________________/
                          exactly zero in value
```

`to_passive` strips every AD layer from a scalar, so each bracket is numerically zero, the forward value is `value` alone, and the derivative is whatever `d_i` says it is. The whole sum is one tape statement whatever the row count. Written the obvious way, `out += d * (x - to_passive(x))` in a loop, it is `n` recorded assignments, and that is how a submodel's entire arithmetic ends up on its consumer's tape one row at a time.

Counted in statements walked, with `T` submodel statements, `k` seeds and `m` outputs:

| | walks per solve |
|---|---|
| recorded inline on the consumer's tape | `T + kT` |
| supplied, nothing recorded | `m + km` |

The second carries no `T`: what a supplied row costs does not depend on how long the submodel is. Rows are checked for finiteness before any is recorded and the report is `[[nodiscard]]`; it is all or nothing, because a value carrying some of its rows is a channel that has gone missing with every number still finite, which is worse than carrying none.

`implicit_value(y_star, dFdy, F)` specialises this to a scalar root: given the root of `F(y, p) = 0` and the residual's slope in the unknown, it records the residual to obtain its slopes in the parameters and yields `dy*/dp = -(dF/dp) / (dF/dy)`.

## The two scalars

`tangent_scalar<T>` (`tangent.hpp`) carries a directional derivative forward and touches no tape. It is what the implicit stepper's Jacobian is taken with. `active_scalar<T>` (`ode_interface.hpp`) is the adjoint scalar a sweep rebinds a System to.

Nesting a tangent above an adjoint is a compile error. At an active inner scalar every operand copy inside an expression template becomes a recorded statement, and because expression templates nest, the growth is superlinear in expression depth: three kernels costing 31 statements written flat cost 566 nested. A slope wanted at an adjoint scalar is taken through the kernel at `double` and handed over as a supplied row.

`value_with_slope<T>` pairs a value with its slope in one type, for a quantity carried with its own derivative (a knot on an interpolated field, a coordinate at an operating point). It declares `for_each_active`, which is why it lives in `odelia`: a bare pair of scalars is a shape the sweep's walk does not open, and would contribute no rows, silently.

## Making a System sweepable

A System that only needs solving provides `ode_size()`, `set_ode_state()`, `ode_state()` and `ode_rates()`, and optionally `ode_time()`, `ode_state_valid()` and `ode_jacobian()`. The compiler-checked parts of that are the concepts `HasOdeTime`, `Rebindable`, `SolvesForValues` and `ChecksState` in `ode_interface.hpp`.

A System a sweep walks adds four members, stated as the concept `Sweepable` there, which `solve_adjoint` asserts:

- `rebind_from<U>()`: a copy of this System on scalar `U`, values only. The sweep builds the adjoint-scalar copy with it.
- `ad_parameters()`: pointers to the parameters the sweep accumulates adjoints for, in a fixed order. The rates must read each one live; a quantity the constructor derived from a parameter carries no adjoint, and that parameter's column comes back zero with no error.
- `for_each_active(f)`: every member carrying the scalar. The sweep hands their tape slots back before it clears the tape; a member left out contributes nothing to the gradient, with every number finite.
- `set_recorded_state(y, time)`: stand on a recorded state, taking the width the recorded time implies.

A System whose state grows adds `apply_insertion(time, x, out)`, the widening as a map, and a System that solves for values inside a stage declares `solved_values` (`SolvesForValues`). The shipped Lorenz example marks which of its members belong to which tier; the smallest complete sweepable System in the tree is `Grow` in `tests/standalone/r_free.cpp`.

## From R: what `$fit()` does

`Lorenz_Solver$set_target(times, target, obs_indices)` stores a schedule and observations; `$fit(ic, params)` returns a least-squares loss and its exact gradient. Underneath (`solver_interface.hpp`, `Solver_fit_impl`): a program of steps is built from `times`, replayed with `advance_recorded` under `set_keep_states(true)`, one seed row of `2 * (y - target)` is placed at each observed index, and `solve_adjoint` is called once per observation interval from the last to the first. `times` must therefore be a step schedule, normally a reference run's `$times()`, not an output grid; a replay that cannot take the schedule's steps one row each is refused. The optimiser is the caller's; `vignette("parameter-fitting")` hands the gradient to `optim()`.

## Reading order through the headers

1. `tangent.hpp` and `value_with_slope.hpp`: the two scalars and the pair. 130 lines, everything uses them.
2. `ode_interface.hpp`: the System concepts, `instruction` and `step_record`, `Sweepable`, `visit_active`.
3. `ode_solver_internal.hpp`: `push_step` and `push_insertion`, the two places a row is written; `step_to` for the pinned path.
4. `adjoint.hpp`: `adjoint_rows`, `active_system`, `vector_jacobian_product`, `state_and_parameter_adjoints`, the transpose of one map.
5. `ode_step_rkck.hpp`: `Step::step_adjoint`, the transpose of one step, and the `solved` channel.
6. `ode_solver.hpp`: `solve_adjoint`, the loop over rows and insertions; `advance_recorded`, the replay.
7. `implicit_node.hpp`: supplied rows and `implicit_value`, independent of the solver.
8. `sweep.hpp`: helpers for a consumer that drives a sweep range by range itself.
